Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-06-22 Manuscript/00-main.Rmd
Improved the the itnroduction part

---
title: Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data
author:
- name: Boyi Guo*
  num: a
- name: Nengjun Yi
  num: a
address:
- num: a
  org: Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA
# - num: b
  # org: Department of Applied Things, University B, City B, Country B
corres: "*Corresponding author name, This is sample corresponding address. \\email{xxx@uab}"
presentaddress: This is sample for present address text this is sample for present address text
authormark: Guo and Yi
articletype: Research article
received: 2021-01-01
revised: 2021-02-01
accepted: 2021-03-01
abstract: "Spike-and-slab priors have been extensively studied for variable selection, and recently been extended to function selection in generalized additive models (GAMs). Most of the previous works on fitting spike-and-slab GAM relies on Markov chain Monte Carlo (MCMC) algorithms to identify the promising subsets of functions. The computational burden of MCMC makes these models less feasible, especially in high-dimensional setting. Besides, most of the current spike-and-slab GAM consider an “all-in-all-out” approach for function selection, rendering less flexibility to answer if non-linear effects are necessary in the model. We propose Bayesian hierarchical generalized additive models for function selection and prediction. The proposed models employ spike-and-slab priors, mixture normal and mixture double exponential, for smooth and sparse solution. Two EM based algorithms, EM-IWLS and EM-Coordinate Descent, are developed for accelerated computation. Simulation studies and real data analysis demonstrated improved performance against the state-of-art models, mgcv and COSSO. The software implementation of the proposed methods is freely available via the R package BHAM."
keywords: Spike-and-Slab Priors; High-Dimensional Data; Generalized Additive Models; EM-IWLS; EM-Coordinate Decent
bibliography: bibfile.bib
output: 
  rticles::sim_article:
    keep_tex: TRUE
---

@@

# References

# Appendix

>>>>

---
title: Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data
author:
- name: Boyi Guo*
  num: a
- name: Byron C. Jaeger
  num: a 
- name: AKM Fazlur Rahman
  num: a   
- name: D. Leann Long
  num: a  
- name: Nengjun Yi
  num: a
address:
- num: a
  org: Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA
# - num: b
  # org: Department of Applied Things, University B, City B, Country B
corres: "*Corresponding author name, This is sample corresponding address. \\email{xxx@uab}"
presentaddress: This is sample for present address text this is sample for present address text
authormark: Guo et al.
articletype: Research article
received: 2021-01-01
revised: 2021-02-01
accepted: 2021-03-01
abstract: "Spike-and-slab priors have been extensively studied for variable selection, and recently been extended to function selection in generalized additive models (GAMs). Most of the previous works on fitting spike-and-slab GAM relies on Markov chain Monte Carlo (MCMC) algorithms to identify the promising subsets of functions. The computational burden of MCMC makes these models less feasible, especially in high-dimensional setting. Besides, most of the current spike-and-slab GAM consider an “all-in-all-out” approach for function selection, rendering less flexibility to answer if non-linear effects are necessary in the model. We propose Bayesian hierarchical generalized additive models for function selection and prediction. The proposed models employ spike-and-slab priors, mixture normal and mixture double exponential, for smooth and sparse solution. Two EM based algorithms, EM-IWLS and EM-Coordinate Descent, are developed for accelerated computation. Simulation studies and real data analysis demonstrated improved performance against the state-of-art models, mgcv and COSSO. The software implementation of the proposed methods is freely available via the R package BHAM."
keywords: Spike-and-Slab Priors; High-Dimensional Data; Generalized Additive Models; EM-IWLS; EM-Coordinate Decent
bibliography: bibfile.bib
output: 
  rticles::sim_article:
    keep_tex: TRUE
---

@@

### 2021-06-22 Manuscript/01-Intro.Rmd
Improved the the itnroduction part

Using additive models to interpolate nonlinear effects has a long history. Early work of additive models can date back to 1979 when @Cleveland1979 proposed local smoothing for bivariate analysis. Since then, many mathematical and statistical models have been proposed, see @Wood2020. With the recent advancement of computational technology, many machine learning algorithms that are computationally intensive start to gain much popularity in prognostic modeling and functional selection, especially when complex signals, for example nonlinear effects, are assumed. Such models generally have improved prediction accuracy over classic statistical models. However, these models have a fatal flaw, the lack of transparency and interpretability, and hence are referred to as "black box" models. Given that the primary objective of biomedical studies is to understand the etiology of diseases, "Black box" models are less helpful as they can hardly explain how the change of biomarkers is associated with the disease.

@@

Many applications of modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There exists an extensive literature on high-dimensional linear models, via penalized models or Bayesian hierarchical models, see @Mallick2013. [TODO a sentence of sparsity. What it means in a penalized model and Bayesian model] Directly apply to additive models, spline models specifically. Problem, oversmoothing of the spline function. An more appropriate treatment would be incorporating the smooth penalty of spline functions.[TODO: a sentence about smoothness penalty]

>>>>

Using additive models to interpolate nonlinear effects has a long history. Early work of additive models can date back to 1979 when \citet{Cleveland1979} proposed local smoothing for bivariate analysis. Since then, many mathematical and statistical models have been proposed, see @Wood2020. With the recent advancement of computational technology, many machine learning algorithms that are computationally intensive start to gain much popularity in prognostic modeling and functional selection, especially when complex signals, for example nonlinear effects, are assumed. Such models generally have improved prediction accuracy over classic statistical models. However, these models have a fatal flaw, the lack of transparency and interpretability, and hence are referred to as "black box" models. Given that the primary objective of biomedical studies is to understand the etiology of diseases, "Black box" models are less helpful as they can hardly explain how the change of biomarkers is associated with the disease.

@@

Many applications of modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see @Mallick2013. These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients: @Ravikumar2009 extended the grouped lasso [@Yuan2006] to additive models; @Huang2010 further developed adaptive grouped lasso for additive models; @Wang2007 and @Xue2009 applied grouped SCAD penalty [@Fan2001] to the additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive MCMC algorithms [@Xu2015; @Yang2020] are proposed where the application on AMs are by-products. @Bai2020 was the first to apply group spike-and-slab lasso prior to AMs using a fast optimization algorithm, and further generalized the framework to GAMs in @Bai2021. Regardless of the nice statistical properties, a practical and pivotal question, if non-linear effects exist, yet to be addressed by these methods. In addition, a common criticism received by these methods is that the basis function coefficients tend to be overly penalized due to the sparsity assumption directly applied on the basis function coefficients. [@Scheipl2013]

To address this problem.

where majority of the efforts ware focusing on the

[TODO a sentence of sparsity. What it means in a penalized model and Bayesian model] Directly apply to additive models, spline models specifically. Problem, oversmoothing of the spline function. An more appropriate treatment would be incorporating the smooth penalty of spline functions.[TODO: a sentence about smoothness penalty]

### 2021-06-22 Manuscript/SS_GAM.tex
Improved the the itnroduction part

\author[a]{Boyi Guo*}
\author[a]{D. Leann Long}
\author[a]{Byron C. Jaeger}
\author[a]{AKM Fazlur Rahman}
\author[a]{Nengjun Yi}

\authormark{Guo and Yi}

@@

In many area of modern biomedical research, dealing with
high-dimensional data, also known as \(p >> n\) problem, is common, for
example sequencing data analysis, electric health record data analysis.
Many statistical models have been proposed for such problem, for
example, penalized models and Bayesian hierarchical models. However, up
to this point, most efforts of methodology development focus on modeling
linear effect in high dimensional setting. It is still not clear what is
the best practice on how to model non-linear effects in high-dimensional
setting.

@@

\hypertarget{references}{%
\section{References}\label{references}}

>>>>

\author[a]{Boyi Guo*}
\author[a]{Byron C. Jaeger}
\author[a]{AKM Fazlur Rahman}
\author[a]{D. Leann Long}
\author[a]{Nengjun Yi}

\authormark{Guo et al.}

@@

Many applications of modern biomedical research, e.g.~sequencing data
analysis, electric health record data analysis, require special
treatment of high-dimensionality, commonly known as \(p >> n\) problem.
There is extensive literature on high-dimensional linear models via
penalized models or Bayesian hierarchical models, see
\citet{Mallick2013}. These models assume sparsity among the predictors,
i.e.~most predictors have small or no effects on the outcome. Further
development focuses on addressing the grouping structure among the
predictors, where AMs and GAMs can be considered special cases. Various
penalties are imposed on the basis function coefficients:
\citet{Ravikumar2009} extended the grouped lasso \citep{Yuan2006} to
additive models; \citet{Huang2010} further developed adaptive grouped
lasso for additive models; \citet{Wang2007} and \citet{Xue2009} applied
grouped SCAD penalty \citep{Fan2001} to the additive models. Recent
Bayesian hierarchical models are also used in the context of
high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive MCMC algorithms
\citep{Xu2015, Yang2020} are proposed where the application on AMs are
by-products. \citet{Bai2020} was the first to apply group spike-and-slab
lasso prior to AMs using a fast optimization algorithm, and further
generalized the framework to GAMs in \citet{Bai2021}. Regardless of the
nice statistical properties, a practical and pivotal question, if
non-linear effects exist, yet to be addressed by these methods. In
addition, a common criticism received by these methods is that the basis
function coefficients tend to be overly penalized due to the sparsity
assumption directly applied on the basis function coefficients.
\citep{Scheipl2013}

To address this problem.

where majority of the efforts ware focusing on the

{[}TODO a sentence of sparsity. What it means in a penalized model and
Bayesian model{]} Directly apply to additive models, spline models
specifically. Problem, oversmoothing of the spline function. An more
appropriate treatment would be incorporating the smooth penalty of
spline functions.{[}TODO: a sentence about smoothness penalty{]}

What is the smooth penalty of spline function?

Many statistical models have been proposed for such problem, for
example, penalized models and Bayesian hierarchical models. However, up
to this point, most efforts of methodology development focus on modeling
linear effect in high dimensional setting. It is still not clear what is
the best practice on how to model non-linear effects in high-dimensional
setting.

@@
